Supplemental Materials

### 2024-07-24 code/analysis-code/fullanalysis-v1.qmd
Final Draft Updates
After meeting with Swartzendruber

# Create a gtsummary table directly from the model
table_model1 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model1

@@

# Create a gtsummary table directly from the model
table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model2

>>>>

# Create a gtsummary table directly from the model
table_model1 <-
  tbl_regression(ML_fit1, exponentiate = FALSE, intercept = TRUE) %>% 
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model1

@@

# Create a gtsummary table directly from the model
table_model2 <- tbl_regression(ML_fit2, exponentiate = FALSE, intercept = TRUE) %>%
  modify_column_unhide(columns = c(statistic, std.error)) %>%
  add_significance_stars(hide_ci = TRUE, hide_p = TRUE) %>%
  add_glance_table(include = c(r.squared, nobs))
table_model2

### 2024-07-24 code/analysis-code/introanalysis-v1.qmd
Draft 1 for advisor
Final changes for draft to advisor
Final Draft Updates
After meeting with Swartzendruber

```{r}
set.seed(123)
ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

@@

# Create a gtsummary table directly from the model
table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) # use exponentiate true for ORs
table_log1

@@

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

@@

# Create a gtsummary table directly from the model
table_log2 <- tbl_regression(log2_fitted, exponentiate = TRUE) # use exponentiate true for ORs
table_log2

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_normalize(all_predictors())

@@

>>>>

```{r}
#set seed
set.seed(123)
ML_log_split <- initial_split(ML_linear, prop = 0.75, strata = BirthweightCategory)

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log1_recipe <- recipe(BirthweightCategory ~ ModifiedColor, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

@@

# Create a gtsummary table directly from the model
table_log1 <- tbl_regression(log1_fitted, exponentiate = TRUE) %>% # use exponentiate true for ORs
  bold_p(t = 0.05) %>%
  modify_table_styling(
    columns = label,
    rows = label == "ModifiedColor",
    text_format = "bold"
  )
table_log1

@@

```{r}
# Simple logistic regression model with Color as the predictor
# Check factor levels
levels(ML_linear$BirthweightCategory) # Reference is LBW

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log2_recipe <- recipe(BirthweightCategory ~ Color, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

@@

# Create a gtsummary table directly from the model
table_log2 <- tbl_regression(log2_fitted, exponentiate = TRUE) %>% # use exponentiate true for ORs
  bold_p(t = 0.05) %>%
  modify_table_styling(
    columns = label,
    rows = label == "Color",
    text_format = "bold"
  )
table_log2

@@

# Create recipe
# Create recipe simple logistic model (creating dummy variables and standardizing continuous variables)
log3_recipe <- recipe(BirthweightCategory ~ ModifiedColor + Age + ModifiedStatus + ModifiedNationality, data = ML_linear) #%>%
  #step_dummy(all_nominal(), -all_outcomes()) %>%
  #step_normalize(all_predictors())

@@

# Merge unadjusted log models 1 and 2

```{r}
# Combine the two tables using tbl_stack for unadjusted ORs
table5 <- tbl_stack(list(table_log1, table_log2), group_header = c()) %>%
  as_gt() %>%
  gt::tab_style(style = gt::cell_text(weight = "bold"),
  locations = gt::cells_row_groups(groups = everything()))
table5

# Save regression output into table
saveRDS(table5, file = here("results", "tables", "table5_final.rds"))
```
```

### 2024-07-24 code/eda-code/eda-v1.qmd
Final Draft Updates
After meeting with Swartzendruber

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))
```

@@

# Create gtsummary table for final analysis
table1 <- 
  ML_summary %>%
  mutate(
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
  ) %>%
  tbl_summary(
    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
    missing = "no",
    label = list(
      ModifiedColor ~ "Ancestry", 
      ModifiedStatus ~ "Parity", 
      Age ~ "Maternal Age", 
      ModifiedNationality ~ "Combined Nationality", 
      Birth ~ "Birth Outcome", 
      MaternalOutcome ~ "Maternal Outcome", 
      FetalOutcome ~ "Fetal Outcome", 
      Lengthcentimeters ~ "Birth Length (cms)", 
      Weightgrams ~ "Birth Weight (grms)",
      WeightgramsBlack ~ "Black (grms)",
      WeightgramsMixed ~ "Mixed-Race (grms)",
      WeightgramsWhite ~ "White (grms)",
      WeightgramsFemale ~ "Female (grms)",
      WeightgramsMale ~ "Male (grms)"
    ),
    statistic = list(
       all_continuous() ~ "{mean} ({sd})",
      WeightgramsBlack ~ "{mean} ({sd})",
      WeightgramsMixed ~ "{mean} ({sd})",
      WeightgramsWhite ~ "{mean} ({sd})",
      WeightgramsFemale ~ "{mean} ({sd})",
      WeightgramsMale ~ "{mean} ({sd})"
    ),
    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
  ) %>%
  add_n() %>%
  modify_table_body( #indent these three rows
    ~.x %>%
      mutate(label = case_when(
        variable == "WeightgramsBlack" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsWhite" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        TRUE ~ label
      ))
  )

>>>>

# Save new summary data with recoded factor variables
saveRDS(ML_summary, file = here("data", "processed-data", "ML_summary.rds"))

#Summarize missing data using dplyr
missing_data_summary <- ML_summary %>% dplyr::summarise_all(list(~sum(is.na(.))))
missing_data_summary
```

@@

# Create gtsummary table for final analysis
table1 <- 
  ML_summary %>%
  mutate(
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
  ) %>%
  tbl_summary(
    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale),
    missing = "no",
    label = list(
      Color ~ "Color (n = 2,695)",
      ModifiedColor ~ "Ancestry (n = 2,695)", 
      ModifiedStatus ~ "Parity (n = 2,836)", 
      Age ~ "Maternal Age (n = 2,783)", 
      ModifiedNationality ~ "Combined Nationality (n = 2,773)", 
      Birth ~ "Birth Outcome (n = 2,761)", 
      MaternalOutcome ~ "Maternal Outcome (n = 2,829)", 
      FetalOutcome ~ "Fetal Outcome (n = 2,666)", 
      Sex ~ "Sex (n = 2,534)",
      Lengthcentimeters ~ "Birth Length (cm) (n = 2,405)", 
      Weightgrams ~ "Birth Weight (g) (n = 2,384)",
      WeightgramsBlack ~ "Black (g) (n = 643)",
      WeightgramsMixed ~ "Mixed-Race (g) (n = 675)",
      WeightgramsWhite ~ "White (g) (n = 950)",
      WeightgramsFemale ~ "Female (g) (n = 1,074)",
      WeightgramsMale ~ "Male (g) (n = 1,270)"
    ),
    statistic = list(
       all_continuous() ~ "{mean} ({sd})",
      WeightgramsBlack ~ "{mean} ({sd})",
      WeightgramsMixed ~ "{mean} ({sd})",
      WeightgramsWhite ~ "{mean} ({sd})",
      WeightgramsFemale ~ "{mean} ({sd})",
      WeightgramsMale ~ "{mean} ({sd})"
    ),
    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
  ) %>%
  #add_n() #%>%
  modify_table_body( #indent these three rows
    ~.x %>%
      mutate(label = case_when(
        variable == "WeightgramsBlack" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsWhite" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        TRUE ~ label
      ))
  )

### 2024-07-24 products/manuscript/supplement/Supplemental_Materials.qmd
Final Draft Updates
After meeting with Swartzendruber

There are three continuous variables: maternal age, infant birth weight, and infant length. There are nine categorical variables: skin color; ancestry; parity or gravidity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.

In the exploratory analysis, I included recorded spontaneous abortions and all recorded live births. For the statistical models, I excluded both spontaneous abortions and any recorded live births weighing \<500 grams to avoid the inclusion of any possible stillbirths that were misrecorded as live births [@nyarko2013; @wehby2015a; @wehby2017a].

A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021]. The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009]. But other health problems related to maternal undernutrition during pregnancy depend on gestational period. Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].

### Maternal Skin Color

The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category. Please see the `eda-v1.qmd` file for more information on the variable manipulation process.

### Gestational Status

Gestational status was divided into parity and gravidity. Originally, the data had eight categories: nullipara, primipara, secundipara, multipara; and primigravida, secundigravida, trigravida, and multigravida. I combined categories into four final categories: 1) nullipara and primigravida into `nullipara` (or any woman who has never given birth to a viable fetus); 2) primipara into `primipara` (or any woman who has given birth to one viable fetus); 3) secundipara and multipara into `multipara` (or any woman who has given birth to more than one viable fetus); and 4) secundigravida, trigravida, and multigravida into `multigravida` (or any woman who has been pregnant more than one but never given birth to a viable fetus).

### Maternal Nationality

For maternal nationality, I created categories based on individual country (`nationality`) and categories based on region (`modifiednationality`). Recategorization occurred as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).

### Clinical Outcomes

The clinical notes upon which I based this dataset only sporadically included infant demographic information for spontaneous abortions or stillbirths. For spontaneous abortions, stillbirths, and live births, I recorded, when available, infant sex, weight, length. I followed the original clinical categorizations for type of delivery in the following cases: natural, indicating minimal medical intervention; interventionist, indicating medium medical intervention through the use of forceps; and operatory, indicating a cesarean section or embryotomy. However, I recategorized external manipulations including version and Mauriceau (used during breech deliveries) (@eyraud1997), coded as operatory or natural by physicians, as interventionist. I classified clinical outcomes as spontaneous abortion, natural (normal delivery), interventionist (interventionist delivery), and operative (surgical delivery).

@@

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
flex_table <- as_flex_table(table3)
flex_table
```
In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, older mothers are associated with giving birth to infants with higher birth weights.

>>>>

There are three continuous variables: maternal age, infant birth weight, and infant length. There are nine categorical variables: skin color; ancestry; parity; nationality; combined nationality; birth outcome; maternal outcome; fetal outcome; and fetal sex.

@@

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table_fit3.rds"))
flex_table <- as_flex_table(table3)
flex_table
```

In the first and third models, there appears to be an association between maternal skin color and infant birth weight, with Euro-descended women (the reference group) having infants with higher birth weights than Afro-descended women. In the second model, older mothers are associated with giving birth to infants with higher birth weights.
